Data migration under a Changing Schema

### 2023-06-28 2022 Migration/articleMigration.tex
WIP introduction

\begin{abstract}
   Incremental software deployment is meant to reduce the time to deploy changes.
   However, an increment might involve a change in the database schema.
   This poses the challenge to preserve already deployed data.
   The problem is how to preserve the semantics of that data under a changing schema.
   This paper proposes a theory for deploying an incremental change,
   meant to be implemented in a software generator called Ampersand.

   This paper aims to preserve the semantics of data, satisfy elementary business requirements,
   and featuring a zero down-time option after deployment.
   We derive a migration schema from both the existing system and a changed schema that specifies the desired system.
   This work is based on the assumptions that
   the migration is meant to deploy a software increment in production,
   the existing data set may be polluted,
   human interaction may be required,
   data meaning must be preserved,
   the business continues during the migration without interruption,
   and there is a compiler to generate an information system from a given schema.
   The correctness of the migration is the focus of this paper, while efficiency is beyond its scope.
\end{abstract}

@@

   This paper is founded in the belief that generating systems can prevent human programming errors, which contributes to robust software.
   Besides, information systems should be released into production in small increments.
   That yields shorter release cycles, more frequent releases, and thus a more agile software development process%
   \footnote{These effects can be measured in terms of DevOps metrics such as
   deployment frequency,
   reliability of deployments, and
   change failure rate~\cite{DevOps2021}.}.
   By deploying frequently, in small increments, and without down-time,
   users get a system that evolves organically.
   This requires a robust data migration process in which the deployment succeeds without failure.
   Automation is required too, to prevent human errors in the process.
   As a result, we expect that migration engineers can deploy smaller increments more frequently, which contributes to a more agile software development process.

@@

\section{Analysis}
\label{sct:Analysis}
   The purpose of an information system is to make data meaningful to its users.
   The analysis of the problem starts with a formal definition of information systems,
   in which the schema contains all information that makes data meaningful.
   We then define a migration procedure from one information system to another,
   which preserves the meaning of the data as much as possible.

@@

   This work uses the Ampersand compiler~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017} to generate information systems.
   It uses a syntactically sugared form of heterogeneous relation algebra,
   which works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
\footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
   Each rule in Ampersand is a constraint (aka invariant), which the system keeps satisfied as long as that rule lives%
\footnote{The dotted lines in the diagrams show which set of rules is enforced.}.
   So, the user spends no effort to extract constraints from specifications or (even more laborious) from code
   because all constraints are explicitly available as rules in the code.
   This, and the absence of imperative code in an Ampersand script, makes Ampersand a suitable platform on which to implement the theory.
   It also allows us to be explicit about ``preserving the meaning as much as possible''.
   An Ampersand script contains just enough information to generate a complete system,
   which means that a classical database schema (i.e.\ data structure plus semantics) can be extracted from the Ampersand script.
   For the purpose of this paper an Ampersand script is considered equivelant to the schema of the generated system.

   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.

   Complications may arise as a result of data pollution,
   which may require human interventions to resolve.
   Data pollution is distinguished in the following categories:
\begin{itemize}
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that cannot be captured in constraints on the data set,
   such as a street address that has become obsolete because someone has moved without notification.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category is about violations of constraints from the schema.
   Such violations do not occur in our theory because the existing system was generated by Ampersand,
   so the data set satisfies all constraints in the schema.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the desired system.
   In that case, a migration engineer must ensure that the desired system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary in specific cases.
   The third category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the desired system.

   This illustrates that automating data migrations may require user intervention,
   either by a migration engineer or by end-users.
   In our approach, such user interventions are given some time by
   defining two distinct moments:
\begin{enumerate}
   \item the moment of transition (the MoT), i.e. the moment the desired system is made available to users;
   \item the moment of cleanup (the MoC), i.e. the moment that the old software and data are removed.%
\footnote{Backup mechanisms are outside the scope of this paper.}.
\end{enumerate}

@@

   In the following section introduces the definitions required to migrate data from one system to another.

@@

   After the migration engineer has produced and tested a migration script, the Ampersand compiler produces a migration system.
   It includes the schemas of both the existing system and the desired system, besides having a schema of its own.
   This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
\begin{figure}
   \centering
   \includegraphics[width=0.8\textwidth]{figures/During migration.png}
   \caption{During the migration}
   \label{fig:during}
\end{figure}
   Deployment of the migration system (figure~\ref{fig:during}) leaves the existing system intact.
   It enhances the system with the data set needed by the migration system, which includes the data set needed by the new system.
   So, users will notice no difference in the existing system when the migration system is deployed.

   When all rules in the migration script have been satisfied, the migration system is ready to be deployed.
   At this point, the ingress switches traffic from the existing system to the migration system.
   Until this moment, users have been updating data in the existing system, which is continuously being migrated to the new system.
   After the ingress switch is made, users have the new functionality at their disposal and make their changes in the new system.
   If for some reason, there is remaining migration work, the migration system will execute this.
   Since there is no more input in the existing system, the migration system will eventually be consistent.
   At this point, the existing system can be removed, which leaves the migration system in place.

>>>>

\begin{abstract}
   Software generators can help to increase the frequency of releases and their reliability.
   They save on time spent on development and time spent on fixing human-induced mistakes by compiling a specification into a working information system.
   However, many generators do not support data migrations, which are necessary when the schema of a system changes during its operational life.
   As a consequence, developers often migrate data ``by hand'', which slows down the pace of incremental software development.

   To address this problem, this paper proposes a theory for data migrations.
   The problem at large is how to preserve the semantics of that data under a changing schema.
   This paper proposes a theory for deploying an incremental change,
   meant to be implemented in a software generator called Ampersand.

   This paper aims to preserve the semantics of data by satisfying concrete business requirements.
   We derive a migration schema from both the existing system and a changed schema that specifies the desired system.
   This work is based on the assumptions that
   the migration is meant to deploy a software increment in production,
   the existing data set may be polluted,
   human interaction may be required,
   data meaning must be preserved,
   the business continues during the migration without interruption (zero down-time),
   and there is a compiler to generate an information system from a given schema.
   The correctness of the migration is the focus of this paper, while efficiency is outside its scope.
\end{abstract}

@@

   This paper is founded in the belief that generating systems can prevent human programming errors, which contributes to robust software.
   Developers can use their efforts saved by generating systems to decrease the time to release successive increments.
   More frequent data migrations are then to be expected.
   A problem is that software generators typically do not support the data migration process, especially when the schema changes.
   With such generators at hand, we can expect developers to spend a larger portion of their time on data migrations,
   which limits the pace of deploying successive releases%
\footnote{These effects can be measured in terms of DevOps metrics such as
   deployment frequency,
   reliability of deployments, and
   change failure rate~\cite{DevOps2021}.}.

   This paper proposes a theory for data migrations that is meant to be implemented in a software generator called Ampersand~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017}.
   It aims at automating data migrations, so incremental changes can be done more reliably and faster.
   This contributes to a more agile software development process.
   We believe that users will experience a more organic evolution of the system because increments are smaller, occur more frequently and more reliably.
   This requires a robust data migration process in which the deployment succeeds without failure.

@@

\section{Analysis}
\label{sct:Analysis}
\subsection{Information Systems}
   The purpose of an information system is to make data meaningful to its users.
   The analysis of the problem starts with a formal definition of information systems,
   in which the schema contains all information that makes data meaningful.
   We then define a migration procedure from one information system to another,
   which preserves the meaning of the data as much as possible.

@@

   Like most other software generators, Ampersand~\cite{Joosten-JLAMP2018, JoostenRAMiCS2017}
   is a compiler that generates information systems from a script.
   Once a generated system is in production, changes that affect the schema are expensive because developers must migrate data ``by hand''.
   This is an error-prone process which slows down the deployment of successive increments.

   Ampersand uses a syntactically sugared form of heterogeneous relation algebra,
   which works with relations in a way similar to Alloy~\cite{Alloy2006} and allegories%
\footnote{Allegories are a specific type of categories, which the reader does not need to understand for the purpose of this paper}~\cite{Zielinski2013}.
   Each rule in Ampersand is a constraint (aka invariant), which the system keeps satisfied as long as that rule lives%
\footnote{The dotted lines in the diagrams show which set of rules is enforced.}.
   So, developers spend no effort to extract constraints from specifications or (even more laborious) from code
   because all constraints are explicitly available as rules in the code.

   This, and the absence of imperative code in an Ampersand script, makes Ampersand a suitable platform on which to implement the theory.
   It also allows us to be explicit about ``preserving the meaning as much as possible''.
   An Ampersand script contains just enough information to generate a complete system,
   which means that a classical database schema (i.e.\ data structure plus semantics) can be extracted from the Ampersand script.

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.

   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
   while transferring data in a controlled fashion.
   A migration strategy (e.g. case by case, customer by customer, or in batches) must ensure the quality of the migration.
   To assume that this can be done fully automatically is unrealistic in many cases%
\footnote{Note that smaller increments improve the chance that a fully automic migration is possible,
although there will always be cases where human intervention is required.}.
   Manual interventions can be necessary to resolve data pollution, new business rules, or to fix known issues in the old system.
   Consequently, a migration engineer must have the freedom to change the migration script to suit the particulars of the migration.
   Such complications may arise mostly as a result of data pollution.
   We distinguish data pollution in the following categories:
\begin{itemize}
   \item data pollution that violates a constraint in the schema.
   \item data pollution that violates a constraint that is not in the schema.
   \item data pollution that cannot be captured in constraints on the data set,
   such as a street address that has become obsolete because someone has moved without notification.
   \item data pollution that is a consequence of an erroneous constraint in the schema.
\end{itemize}
   The first category is about violations of constraints from the schema.
   Such violations do not occur in our theory because the existing system was generated by Ampersand,
   so the data set satisfies all constraints in the schema.
   This sets our approach apart from other approaches to formalize data migration, e.g.~\cite{Thalheim2013}.
   The second category, data pollution that is not captured by any of the constraints mentioned in the schema,
   can be dealt with by adding constraints in the desired system.
   In that case, a migration engineer must ensure that the desired system can start with data that satisfies all constraints in its schema.
   So some human intervention may be necessary in specific cases.
   The third category of pollution must be dealt with by working procedures, to prevent such pollution as much as possible.
   In some cases, constraints on data can be formulated,
   so the system can help the user and simplify (or eliminate) the corresponding working procedure.
   So, such constraints may help to move this type of pollution to one of the other categories.
   The last category of data pollution, erroneous constraints, must be fixed by replacing them by the correct constraints in the desired system.

   This illustrates that automating data migrations may require user intervention,
   either by a migration engineer or by end-users.
   In our approach, such user interventions are given some time by
   defining two distinct moments:
\begin{enumerate}
   \item the moment of transition (the MoT), i.e. the moment that changed functionality is made available to users;
   \item the moment of cleanup (the MoC), i.e. the moment after which the old software and data can be removed.%
\footnote{Backup mechanisms are outside the scope of this paper.}.
\end{enumerate}

@@

   The following section introduces the definitions required to migrate data from one system to another.

@@

   After the migration engineer has produced and tested a migration script, the Ampersand compiler produces a migration system.
   It includes the schemas of both the existing system and the desired system, besides having a schema of its own.
   This is implemented by Ampersand's INCLUDE mechanism, which boils down to taking the disjoint union of the schemas.
\begin{figure}
   \centering
   \includegraphics[width=0.8\textwidth]{figures/During migration.png}
   \caption{During the migration}
   \label{fig:during}
\end{figure}
   Deployment of the migration system (figure~\ref{fig:during}) leaves the existing system intact.
   It enhances the system with the data set needed by the migration system, which includes the data set needed by the desired system.
   So, users will notice no difference in the existing system when the migration system is deployed.

   When all rules in the migration script have been satisfied, the migration system is ready to be deployed.
   At this point, the ingress switches traffic from the existing system to the migration system.
   Until this moment, users have been updating data in the existing system, which is continuously being migrated to the desired system.
   After the ingress switch is made, users have the new functionality at their disposal and make their changes in the desired system.
   If for some reason, there is remaining migration work, the migration system will execute this.
   Since there is no more input in the existing system, the migration system will eventually be consistent.
   At this point, the existing system can be removed, which leaves the migration system in place.
